Equalizer Programming Guide

### 2008-12-01 Developer/ProgrammingGuide/paper.tex
0.6 release updates
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.17 for Equalizer 0.6-RC1\\[\medskipamount]
  \today
}

@@

\maketitle
\vfill
\lstset{language=C++}

\pagestyle{empty}
\maketitle
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock}

@@

\section*{Equalizer 0.6 Programming Guide}
\subsection*{Contributors}

@@

The images on the front page show the following Equalizer applications:
a terrain rendering application on a six-node display wall [top left], a
flow visualization application for climate research\footnote{Image
  courtesy of Computer Graphics and Multimedia Systems, University of
  Siegen}, the \textsf{eVolve} volume renderer\footnote{Data set
  courtesy of General Electric, USA} [bottom left] the \textsf{eqPly}
polygonal renderer in a three-sided CAVE [bottom center], and a six-node
sort-last database decomposition with parallel direct-send
recomposition\footnote{Data set courtesy of Stanford University Computer
  Graphics Laboratory} [bottom right].

@@

\clearpage
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
    1.3.17  & Nov 13, 2008 & Added \sref{sLoadBalancing} and \sref{sAdvDPlex}, final touches\\
    \hline
  \end{tabularx}}
\clearpage

@@

@@

In order to adapt an application for Equalizer, the programmer
structures the source code so that the OpenGL rendering can be executed
in parallel, potentially using multiple processes for cluster-based
execution. Equalizer provides the domain-specific parallel rendering
know-how and abstracts configuration, threading, synchronization,
windowing and event handling. It is a `GLUT on steroids', providing
parallel and distributed execution, scalable rendering features and
fully customizable event handling.

@@

\section{Getting Started}

@@

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and compiling the source code. After installing Equalizer, please take a
look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of the \textsf{eqPly} example.

Compiling Equalizer is as simple as running \textsf{make} on Linux and
Mac OS X or building the Equalizer Visual Studio 2005 solution on
Windows. Note that on Mac OS X 10.4 (Tiger), some prerequisites have to
be installed before running \textsf{make}, as explained in
\textsf{README.Darwin}.

@@

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
only supports one application per server, but it will provide
concurrent and efficient multi-application support in future.

@@

The application connects to an Equalizer server and receives a
configuration.  Furthermore, the application also provides its render
client, which will be controlled by the server. The application reacts
on events, updates its database and controls the rendering.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client-related code, it is reduced to be the application's `master'
process without any OpenGL windows and rendering code.

The rendering client can be the same executable as the application, as
it is the case with all provided examples. When it is started as a
render client, the Equalizer initialization routine does not return and
takes over the control by calling the render client task
methods. Complex applications usually implement a separate, light-weight
rendering client.

\subsubsection{Parallel Rendering}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
\end{wrapfigure}

@@

The \textsf{eqHello} redraw function renders six colored quads, rotating
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convience function to setup the
frustum and other OpenGL state. After setting up some lighting
parameters, \textsf{eqHello} rotates the scene and renders the quads
using immediate mode:

@@

Real-time visualization is an inherently parallel
problem. Unfortunately, different applications have different rendering
algorithms, which require different scalable rendering modes to address
the right bottlenecks. Equalizer supports the most important algorithms,
and will continue to add new ones over time to meet application
requirements.

This section gives an introduction to scalable rendering, in order to
provide some background for application developers. The scalability
modes offered by Equalizer are discussed, along with their advantages
and disadvantages.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\small A database compound}}
\end{wrapfigure}
DB\footnote{3D model courtesy of AVS, USA.} decomposes the
rendered database so that all rendering units process a part of the
scene in parallel. This mode is also known as sort-last.

@@

The application has to partion the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

The best use of database compounds is to divide the data to a managable
size, and then to use other decomposition modes to achieve further
scalability. Volume rendering is one of the applications which can
profit from database compounds.

DB compounds in Equalizer are configured using the \textsf{range}
parameter, using the values \textsf{[ begin end ]} in normalized
coordinates. The range defines the start and end point of the
application's database to be rendered. The value has to be interpreted
by the application's rendering code accordingly.  Each child compound
uses an output frame, which is connected to an input frame on the
destination channel. For more than two contributing channels, it is
recommended to configure parallel direct send compositing, as described
in \sref{sDirectSend}.

@@

Stereo compounds\footnote{3D model courtesy of Stereolithography Archive
  at Clemson University.} assign each eye pass to individual rendering
units. The resulting images are copied to the appropriate stereo
buffer. This mode supports a variety of stereo modes, including active
(quad-buffered) stereo, anaglyphic stereo and auto-stereo displays with
multiple eye passes.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\small A Stereo compound}}
\end{wrapfigure}
Due to the frame consistency between the eye views, this modes scales
very well. The IO requirements for pixel transfer are small and
constant.

@@

DPlex compounds\footnote{3D model courtesy of Cyberware, Inc.} assign
full, alternating frames to individual rendering units. The resulting
images are copied to the destination channel, and Equalizer
load-balancing is used to ensure a steady framerate on the destination
window. This mode is also known as time-multiplex or AFR.

@@

Pixel compounds are similar to 2D compounds. The frustra of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\small A pixel compound}}
\end{wrapfigure}
Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources.

@@

Pixel compounds are ideal for raytracing, which is highly fill-limited
and needs the full database for rendering anyway. Volume rendering
applications are also well suited for this mode, and should choose it
over 2D compounds.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small Static 2D compound vs 2D
      load-balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D, DB and
DPlex compounds might need load-balancing to provide optimal
performance.

Equalizer provides automatic, application-transparent load-balancing for
2D and DB compounds. Using an appropriate load-balancer for these
compounds will adapt the 2D split or database range each frame. The
current load-balancing algorithm is a technology preview and will
improve in future versions.

2D load-balancing gives better framerates than static decomposition in
virtually all cases. DB load-balancing is benefitial for applications
which can not precisely predict the load for their scene data. Volume
rendering is a counter-example, where the data is uniform and a static
DB decomposition typically results in a better performance.

DPlex compounds require a smoothing of the framerate at the destination
channel, otherwise the framerate will become periodically faster and
slower. Using a DPlex load-balancer will smoothen the swap buffer rate
on the destination channel for optimal user experience.

@@

The application inherits from Equalizer classes and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are similar in concept to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

@@

All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the scene data has to be maintained within a process
since all rendering threads are guaranteed to draw the same frame.

@@

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread, or
in the case of non-threaded pipes from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. Framebuffer object (FBO) support will be implement by a later
version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
abstracts and manages window-system-specific handles to the drawable and
context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
the glX window system.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

The frustrum can be specified as a wall or projection description.

@@

\fig{fFrustra} illustrates the wall and projection frustum parameters.

\subsubsection{Compound Classification}
The channels of the leaf compounds in the compound tree are designated
as source channels. The top-most channel in the tree is the destination
channel. Only source channels execute the draw task. All channels in a
compound tree work for the destination channel. The destination channel
defines the 2D pixel viewport rendered by all leaf compounds. The
destination channel and pixel viewport can not be overridden by child
compounds.

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. A \textsf{pixel} parameter adapts the frustum
so that the source channel renders an evenly subset of the parent's
pixels. Setting one or multiple attributes causes the parent's view to
be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

\subsubsection{Recomposition - Frames}
Compounds use output and input frames to configure the recomposition of
the resulting pixel data from the source channels. An output frame
connects to an input frame of the same name. The selected frame buffer
data is transported from the output channel to the input channel. The
assembly routine of the input channel will block on the availability of
the output frame. This composition process is extensively described in
\sref{sCompositing}. Frame names are only valid within the compound
tree, that is, an output frame from one compound tree can not be used as
an input frame of another compound tree.

@@

Note that this software swap barrier is not accurate enough for
edge-blended projection system. Such installations typically use a
hardware synchronization, e.g., nVidia G-Sync cards. It is however
sufficient for display walls made out of LCD's.

Swapbarrier names are only valid within the compound tree, that is, a
compound from one compound tree can not be synchronized with a compound
from another compound tree.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. It focuses on the example usage of
core Equalizer features, not on rendering features or visual quality.

@@

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sConfig}, and two other classes
for distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

The distributed data classes are helper classes based on
\textsf{eq::net::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data.

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig()  
        { return new eqPly::Config; }
    virtual eq::Node*    createNode( eq::Config* parent )  
        { return new eqPly::Node( parent ); }
    virtual eq::Pipe*    createPipe( eq::Node* parent )
        { return new eqPly::Pipe( parent ); }
    virtual eq::Window*  createWindow( eq::Pipe* parent )
        { return new eqPly::Window( parent ); }
    virtual eq::Channel* createChannel( eq::Window* parent )
        { return new eqPly::Channel( parent ); }
};
\end{lstlisting}}

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eq::net::Node}. The underlying network layer in
Equalizer is a peer-to-peer network of \textsf{eq::net::Node}s. The
application programmer does not usually have to be aware of the classes
in the \textsf{eq::net} namespace, but both the \textsf{eq::Client} and
the server are \textsf{eq::net::Node}s.

The local initialization of a node creates a local listening socket,
which allows the \textsf{eq::Client} to communicate over the network
with other nodes, such as the server and the rendering clients.

@@

In the case of \textsf{eqPly}, the application is also the render
client. The \textsf{eqPly} executable has three run-time behaviours:

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity in the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.4 for Equalizer 0.6\\[\medskipamount]
%  \today
}

@@

\pagestyle{empty}
\maketitle
\thispagestyle{empty}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock}

@@

\lstset{language=C++}
\section*{Equalizer 0.6 Programming Guide}
\today
\subsection*{Contributors}

@@

The images on the front page show the following applications build using
Equalizer: RTT DeltaGen\footnote{Image copyright Realtime Technology AG,
  2008} [top left], a flow visualization application for climate
research\footnote{Image courtesy of Computer Graphics and Multimedia
  Systems, University of Siegen}, the \textsf{eVolve} volume
renderer\footnote{Data set courtesy of General Electric, USA} [bottom
left] the \textsf{eqPly} polygonal renderer in a three-sided CAVE
[bottom center], and a six-node sort-last database decomposition with
parallel direct-send recomposition\footnote{Data set courtesy of
  Stanford University Computer Graphics Laboratory} [bottom right].

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer is used for scientific simulation visualization at various
research institutes, for creating electronic art exhibitions as well as
in commercial 3D design applications.

@@

To adapt an application for Equalizer, the programmer structures the
source code so that the OpenGL rendering can be executed in parallel,
potentially using multiple processes for cluster-based execution.

Equalizer provides the domain-specific parallel rendering know-how and
abstracts configuration, threading, synchronization, windowing and event
handling. It is a `GLUT on steroids', providing parallel and distributed
execution, scalable rendering features, network data distribution and
fully customizable event handling.

@@

\subsection{Parallel Rendering}

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}\vspace{-2ex}
\end{wrapfigure}

@@

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and compiling the source code. After installing Equalizer, please take a
look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of Equalizer and the
\textsf{eqPly} example.

Compiling Equalizer is as simple as running \textsf{make} on Linux or
Mac OS X, building the provided XCode project on Mac OS X 10.5, or
building the Equalizer Visual Studio 2005 solution on Windows. Note that
on Mac OS X 10.4 (Tiger), some prerequisites have to be installed before
running \textsf{make}, as explained in \textsf{README.Darwin}.

@@

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
only supports one application per server, but it will provide
concurrent and efficient multi-application support in the future.

@@

The application connects to an Equalizer server and receives a
configuration.  Furthermore, the application also provides its render
client, which will be controlled by the server. The application reacts
on events, updates its data and controls the rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client code, it is reduced to be the application's `master' process
without any OpenGL windows and rendering code.

The rendering client can be the same executable as the application, as
it is the case with all provided examples. When it is started as a
render client, the Equalizer initialization routine does not return and
takes over the control by calling the render client task
methods. Complex applications usually implement a separate, lightweight
rendering client.

@@

The \textsf{eqHello} redraw function renders six colored quads, rotating
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convenience function to setup the
frustum and other OpenGL state. After setting up some lighting
parameters, \textsf{eqHello} rotates the scene and renders the quads
using immediate mode:

@@

Real-time visualization is an inherently parallel
problem. Unfortunately, different applications have different rendering
algorithms, which require different scalable rendering modes to address
the bottlenecks correctly. Equalizer supports all important algorithms as
listed below, and will continue to add new ones over time to meet
application requirements.

This section gives an introduction to scalable rendering, to
provide some background for application developers. The scalability
modes offered by Equalizer are discussed, along with their advantages
and disadvantages.

@@

DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.},
decomposes the rendered database so that all rendering units process a
part of the scene in parallel. This mode is also known as sort-last, and
is very similar to the data decomposition approach used by HPC
applications.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}\small A database compound}}
\end{wrapfigure}
The application has to partition the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

The best use of database compounds is to divide the data to a manageable
size, and then to use other decomposition modes to achieve further
scalability. Volume rendering is one of the applications which can
profit from database compounds.

DB compounds in Equalizer are configured using the \textsf{range}
parameter, using the values \textsf{[ begin end ]} in normalized
coordinates. The range defines the start and end point of the
application's database to be rendered. The value has to be interpreted
by the application's rendering code accordingly.  Each child compound
uses an output frame, which is connected to an input frame on the
destination channel. For more than two contributing channels, it is
recommended to configure streaming or parallel direct send compositing,
as described in \sref{sDirectSend}.

@@

Stereo compounds, as shown in \fig{fStereoCmp}\footnote{3D model
  courtesy of Stereolithography Archive at Clemson University.}, assign
each eye pass to individual rendering units. The resulting images are
copied to the appropriate stereo buffer. This mode supports a variety of
stereo modes, including active (quad-buffered) stereo, anaglyphic stereo
and auto-stereo displays with multiple eye passes.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo compound}}
\end{wrapfigure}
Due to the frame consistency between the eye views, this modes scales
very well. The IO requirements for pixel transfer are small and
constant.

@@

DPlex compounds assign full, alternating frames to individual rendering
units. The resulting images are copied to the destination channel, and
Equalizer load-balancing is used to ensure a steady framerate on the
destination window. This mode is also known as time-multiplex or AFR.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}\vspace{-5ex}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. The frusta of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels, as shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}

@@

@@

Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources. Pixel compounds are ideal for raytracing,
which is highly fill-limited and needs the full database for rendering
anyway. Volume rendering applications are also well suited for this
mode, and should choose it over 2D compounds.

@@

While pixel and stereo compounds are naturally load-balanced, 2D, DB and
DPlex compounds might need load-balancing to provide optimal
performance. Using a load-balancer is transparent to the application,
and can be used on any application for 2D and DPlex, and on most
applications for DB load-balancing. Some applications supporting DB
decomposition do not support dynamic updates of the database range, and
therefore cannot be used with DB load-balancing.

Using a 2D or DB load-balancer will adapt the 2D split or database range
automatically each frame. The 2D load-balancer exists in three flavors:
\textsf{2D} using tiles, \textsf{VERTICAL} using columns and
\textsf{HORIZONTAL} using rows. The current load-balancing algorithm for
2D and DB is a technology preview and will improve in future versions.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
\end{wrapfigure}
2D load-balancing increases the framerate over a static decomposition in
virtually all cases. It works best if the application data is relatively
uniformly distributed in screen-space.

DB load-balancing is beneficial for applications which cannot precisely
predict the load for their scene data. Volume rendering is a
counterexample, where the data is uniform and a static DB decomposition
typically results in a better performance.

DPlex compounds require a smoothing of the framerate at the destination
channel, otherwise the framerate will become periodically faster and
slower. Using a DPlex load-balancer will smoothen the swap buffer rate
on the destination window for optimal user experience.

@@

The application inherits from Equalizer classes and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are similar in concept to C function callbacks. \sref{sEqPly}
will discuss the important task methods. A full list can be found on the
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchronization or protection of shared data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

Rendering threads on a single node are synchronized when using the
default thread model \textsf{draw\_sync}. When a frame is finished, all
local rendering threads are done drawing. Therefore the application can
safely modify the data between the end of a frame and the beginning of a
new frame. Furthermore, only one instance of the scene data has to be
maintained within a process, since all rendering threads are guaranteed
to draw the same frame.

@@

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. They
should only be used if using a different, but synchronized rendering
thread is not an option.

All pipe, window and channel task methods are executed from the pipe
thread, or in the case of non-threaded pipes from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. FrameBuffer object (FBO) support will be implement by a later
version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
abstracts and manages window-system-specific handles to the drawable and
context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
the glX window system.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\small\label{fFrusta}Wall and projection parameters}}
\end{wrapfigure}
The frustum description is inherited by the children, therefore the
frustum is typically defined on the topmost compound.

The frustum can be specified as a wall or projection description.

@@

\fig{fFrusta} illustrates the wall and projection frustum parameters.

\subsubsection{Compound Classification}
The channels of the leaf compounds in the compound tree are designated
as source channels. The topmost channel in the tree is the destination
channel. One compound tree might have multiple destination channels,
e.g., for a swap-synchronized immersive installation.

All channels in a compound tree work for the destination channel. The
destination channel defines the 2D pixel viewport rendered by all leaf
compounds. The destination channel and pixel viewport cannot be
overridden by child compounds.

\subsubsection{Tasks}
Compounds execute a number of tasks: clear, draw, assemble and
readback. By default, a leaf compound executes all tasks and a non-leaf
compound assemble and readback. A non-leaf compound never executes the
draw task.

A compound can be configured to execute a specific set of tasks, for
example to configure the multiple steps used by binary-swap compositing.

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adapts the frustum so that the source channel
renders an evenly subset of the parent's pixels. Setting one or multiple
attributes causes the parent's view to be decomposed
accordingly. Attributes are cumulative, that is, intermediate compound
attributes affect and therefore decompose the rendering of all their
children.

\subsubsection{Recomposition - Frames}
Compounds use output and input frames to configure the recomposition of
the resulting pixel data from the source channels. An output frame
connects to an input frame of the same name. The selected frame buffer
data is transported from the output channel to the input channel. The
assembly routine of the input channel will block on the availability of
the output frame. This composition process is extensively described in
\sref{sCompositing}. Frame names are only valid within the compound
tree, that is, an output frame from one compound tree cannot be used as
an input frame of another compound tree.

@@

Note that this software swap barrier is not accurate enough for
edge-blended projection system. Such installations typically use a
hardware synchronization, e.g., nVidia G-Sync cards. It is however
sufficient for display walls made out of LCD's and for some
non-immersive projection systems.

Swapbarrier names are only valid within the compound tree, that is, a
compound from one compound tree cannot be synchronized with a compound
from another compound tree.

@@

\textsf{eqPly} is a parallel renderer for polygonal data in the
\textsf{ply} file format. It supports all relevant Equalizer features,
and can be used to render on large-scale displays, immersive
environments with head tracking and to render massive data sets using
all scalable rendering features of Equalizer.

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. It focuses on the example usage of
core Equalizer features, not on advanced rendering features or visual
quality.

@@

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sConfig}, and classes for
distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

The distributed data classes are helper classes based on
\textsf{eq::net::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data. Furthermore
they are used for a basic scene graph distribution of the model data.

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig( eq::ServerPtr parent )
        { return new eqPly::Config( parent ); }
    virtual eq::Node*    createNode( eq::Config* parent )  
        { return new eqPly::Node( parent ); }
    virtual eq::Pipe*    createPipe( eq::Node* parent )
        { return new eqPly::Pipe( parent ); }
    virtual eq::Window*  createWindow( eq::Pipe* parent )
        { return new eqPly::Window( parent ); }
    virtual eq::Channel* createChannel( eq::Window* parent )
        { return new eqPly::Channel( parent ); }
};
\end{lstlisting}}

The third step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
underlying network layer in Equalizer is a peer-to-peer network of
\textsf{eq::net::Node}s. The client/server concept is implement the
higher-level \textsf{eq} client namespace.

The application programmer does not usually have to be aware of the
classes in the \textsf{eq::net} namespace, but both the
\textsf{eq::Client} and the Equalizer server are
\textsf{eq::net::Node}s.

The local initialization of a node creates at least one local listening
socket, which allows the \textsf{eq::Client} to communicate over the
network with other nodes, such as the server and the rendering
clients. The listening socket(s) can be configured using the
\textsf{--eq-listen} command line parameter, or by programmatically
adding connection descriptions to the client before the local
initialization.

@@

In the case of \textsf{eqPly}, the application is also the render
client. The \textsf{eqPly} executable has three run-time behaviors:

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity of the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}. The launch
  command can be used to configure another program to auto-launch render
  clients.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

    // All render data is static or multi-buffered, we can run asynchronously
    if( getIAttribute( IATTR_THREAD_MODEL ) == eq::UNDEFINED )
        setIAttribute( IATTR_THREAD_MODEL, eq::ASYNC );
